Error Reporting Parsers: a Monad Transformer Approach

### 2013-06-24 article.tex
Parsers:  use refurbished code in article examples, and improve explanations

And some examples to make sure it's behaving correctly: \EZY{An example?! Why not a QuickCheck property? \verb|;-)|}
\begin{verbatim}
*Main> map (runParser  opencurly) ["{abc", "}abc", "(abc", "abc"]
[Just ('{', "abc"), Nothing, Nothing, Nothing]
\end{verbatim}

Good, it correctly accepts open curly brackets and rejects everything else.
But what about the spec that said that comments and whitespace can occur 
anywhere before or after tokens -- how should that be handled?

@@

So what we want to throw away are comments and whitespace; and let's throw them 
away after every token, using a combinator which first parses a token, and then
throws away junk:
\begin{verbatim}
junk = many0 (whitespace <|> comment)

@@

Here are these three parsers in action:
\begin{verbatim}
*Main> runParser whitespace "\n \t   \tabc"
Just ("\n \t   \t","abc")

*Main> runParser whitespace "abc"
Nothing

*Main> runParser comment ";123  \t\nabc"
Just (";123  \t","\nabc")

*Main> runParser (munch $ literal 'a') "abc"
Just ('a',"bc")

*Main> runParser (munch $ literal 'a') ";comment!\n   abc"
Just ('a',"bc")

*Main> runParser (munch $ literal 'a') ";comment!\n   bca"
Nothing

*Main> runParser (munch $ literal 'a') "bca"
Nothing
\end{verbatim}

@@

With \verb+munch+ in hand, we fix \verb+opencurly+ and try it out:
\begin{verbatim}
opencurly = munch $ literal '{'

*Main> runParser opencurly ";abc\n{def"
Just ('{', "def")

*Main> runParser opencurly "{def"
Just ('{', "def")

*Main> runParser opencurly "(def"
Nothing

*Main> runParser opencurly "}def"
Nothing
\end{verbatim}

Excellent!  The \verb+opencurly+ parser correctly throws away whitespace 
and comments and succeeds, consuming one character, when the first non-junk 
character is an open curly brace, failing otherwise.

It's straightforward to define parsers for the other three brace tokens:
\begin{verbatim}
closecurly = munch $ literal '}'
openparen  = munch $ literal '('
closeparen = munch $ literal ')'
\end{verbatim}

The parser for our last token, Symbol, is more complicated because any number of
alphabetical characters is valid.  We'll use \verb+satisfy+ and an appropriate
predicate to check that a character is valid, and \verb+many1+ to 
match any positive number of valid characters:
\begin{verbatim}
symbol = munch $ many1 char
  where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))

@@

*Main> runParser symbol ";sdfasdfsa\n  abc123"
Just ("abc","123")

*Main> runParser symbol ";sdfasdfsa\n  123abc"
Nothing
\end{verbatim}

Note that we're again using \verb+munch+ to throw away leading junk.

@@

Next, we move on to \verb+define+.
A straightforward translation from the grammar produces:
\begin{verbatim}
define =
    opencurly                    *>
    check (== "define") symbol   *>
    pure ADefine                <*>
    symbol                      <*>
    form                        <*
    closecurly
\end{verbatim}

On to Lambda!  Not only does Lambda have additional syntax with its parameter 
list, but we also need to ensure that the parameter names are unique.  We can 
do that using the \verb+check+ combinator again:
\begin{verbatim}
lambda = 
    opencurly                       *>
    check (== "lambda") symbol      *>
    opencurly                       *>
    pure ALambda                   <*>
    check distinct (many0 symbol)  <*>
    (closecurly                     *>
     many1 form                    <*
     closecurly)
  where
    distinct names = length names == length (nub names)
\end{verbatim}
If the symbols are distinct, the parameter list subparser will succeed,
whereas if there's a repeated symbol, it will fail.

\subsection{Finishing touches}
All we have left to do are the \verb+special+, \verb+form+ (which 
we've already used to build the previous parsers), and \verb+woof+ parsers.  
The special parser is simple -- it parses either define or lambda:
\begin{verbatim}
special = define <+> lambda
\end{verbatim}

@@

The final parser, \verb+woof+, needs to not only parse all the forms, but also 
make sure that the entire input is consumed.  The end of input check is done 
with the \verb+end+ parser:
\begin{verbatim}
end = switch item
\end{verbatim}

We also need to munch any trailing comments and whitespace -- otherwise \verb+end+
would fail -- so the final Woof parser is:
\begin{verbatim}
woof = many1 form <* munch end
\end{verbatim}

\subsection{Examples}
Let's look at a few examples of our parser in action:
\begin{verbatim}
*Main> runParser woof "; \n   {lambda {x y z} a b (c b a)}end"
Just ([ALambda ["x","y","z"] 
               [ASymbol "a",
                ASymbol "b",
                AApp (ASymbol "c") 
                     [ASymbol "b",
                      ASymbol "a"]],
       ASymbol "end"],
      "")

*Main> runParser woof "a b c {define q (f x)}"
Just ([ASymbol "a",
       ASymbol "b",
       ASymbol "c",
       ADefine "q" (AApp (ASymbol "f") [ASymbol "x"])],"")

*Main> runParser woof "a b c {define q (f x)},"
Nothing
\end{verbatim}
Note that the first two parses consume the entire input, while the last one 
fails because it doesn't recognize the trailing comma.

@@

Let's start with an informal spec of different cases of malformed input, along 
with the error messages that should be reported: 
\begin{itemize}
  \item application: missing operator:  \begin{verbatim}()\end{verbatim}
  \item application: missing close parenthesis:  \begin{verbatim}(a b\end{verbatim}
  \item define: missing symbol:  \begin{verbatim}{define (a b c)}\end{verbatim}
  \item define: missing form:  \begin{verbatim}{define a}\end{verbatim}
  \item define: missing close curly:  \begin{verbatim}{define a b\end{verbatim}
  \item lambda: missing parameter list:  \begin{verbatim}{lambda (a b c)}\end{verbatim}
  \item lambda: duplicate parameter names:  \begin{verbatim}{lambda {a b a} (c d)}\end{verbatim}
  \item lambda: missing parameter list close curly:  \begin{verbatim}{lambda {a b (c d)}\end{verbatim}
  \item lambda: missing body form:  \begin{verbatim}{lambda {a b}}\end{verbatim}
  \item lambda: missing close curly:  \begin{verbatim}{lambda {a b} (c d)\end{verbatim}
  \item special form: unable to parse:  \begin{verbatim}{defin x y}\end{verbatim}
  \item woof: unparsed input:  \begin{verbatim}a,b\end{verbatim}
\end{itemize}
To allow error reporting in our parsers, we first need to extend our monad stack, 
adding in a layer for errors.  We'll use a second monad transformer type class 
-- MonadError, replace the Maybe layer with its corresponding transformer datatype 
MaybeT, and add in an Either layer on the bottom of the stack.  Our parser stack 
and \verb+runParser+ function are now:
\begin{verbatim}
type Parse e t a = StateT [t] (MaybeT (Either e)) a

@@

*Main> runParser (throwError "no!" <|> satisfy (== 'b')) "babc"
Left "no!"
\end{verbatim}
We'll take advantage of this backtracking restriction to produce accurate error
messages.

@@

\subsection{Error messages}
We must modify the remaining parsers to report meaningful error messages; let's
define the messages that will be reported in each case:
\begin{verbatim}
eAppOper    =  "application: missing operator"
eAppClose   =  "application: missing close parenthesis"
eDefSym     =  "define: missing symbol"
eDefForm    =  "define: missing form"
eDefClose   =  "define: missing close curly"
eLamParam   =  "lambda: missing parameter list"
eLamDupe    =  "lambda: duplicate parameter names"
eLamPClose  =  "lambda: missing parameter list close curly"
eLamBody    =  "lambda: missing body form"
eLamClose   =  "lambda: missing close curly"
eSpecial    =  "special form: unable to parse"
eWoof       =  "woof: unparsed input"
\end{verbatim}

@@

What data do we want to include in our errors?  A String describing the error 
and a token reporting the position will do:
\begin{verbatim}
type Error = (String, Token)
\end{verbatim}

Now let's start with \verb+application+ as our first modified, error-reporting 
parser:
\begin{verbatim}
application =
    openparen                     >>
    commit eAppOper form          >>= \op ->
    many0 form                    >>= \args ->
    commit eAppClose closeparen   >>
    return (AApp op args)
\end{verbatim}

@@

The \verb+define+ and \verb+lambda+ parsers are upgraded similarly:
\begin{verbatim}
define =
    check (== "define") symbol    *>
    pure ADefine                 <*>
    commit eDefSym symbol        <*>
    commit eDefForm form         <*
    commit eDefClose closecurly

lambda =
    check (== "lambda") symbol       >>
    commit eLamParam opencurly       >>
    many0 symbol                     >>= \params ->
    (if distinct params 
        then return ()
        else throwError eLamDupe)    >>
    commit eLamPClose closecurly     >>
    commit eLamBody (many1 form)     >>= \bodies ->
    return (ALambda params bodies)   <*
    commit eLamClose closecurly  
  where
    distinct names = length names == length (nub names)
\end{verbatim}

@@

\subsection{Error messages:  finishing touches}
To complete the error-reporting parser, we need to change a couple more details. 
First, we want an error reported when an open curly brace is found, but no 
special form can be parsed.  We extend the \verb+special+ 
parser to report an error if it matches an open brace but not define or lambda:
\begin{verbatim}
special = 
    opencurly  *>  commit eSpecial (define <|> lambda)
\end{verbatim}

@@

When we can no longer parse any forms, instead of failing, we'd like to report 
an error if there's any input left, and otherwise succeed.  Our ending parser
must first munch any leading whitespace or comments, as usual, and then check 
whether there are any characters left; if there are, it should report an error:
\begin{verbatim}
endCheck = 
    many0 (whitespace <|> comment) *>
    switch item
\end{verbatim}
The \verb+woof+ parser uses \verb+endCheck+ after consuming all possible forms:
\begin{verbatim}
woof = many0 form <* commit eWoof endCheck
\end{verbatim}

@@

Knowing that a close-parenthesis is missing, somewhere deep inside a nested mess
of forms isn't very helpful -- you also want to know where it happened.  Let's 
add that to our parser.

>>>>

It accepts open curly brackets and rejects everything else:
\begin{verbatim}
*Main> map (runParser  opencurly) ["{abc", "}abc", "(abc", "abc"]
[Just ('{', "abc"), Nothing, Nothing, Nothing]
\end{verbatim}

But what about the spec that said that comments and whitespace can occur 
anywhere before or after tokens -- how should that be handled?

@@

We want to throw away comments and whitespace, and let's throw them away after 
every token.  Let's do that using a combinator which first parses a token, and 
then throws away junk:
\begin{verbatim}
junk = many0 (whitespace <|> comment)

@@

@@

With \verb+tok+ in hand, we fix \verb+opencurly+ and try it out:
\begin{verbatim}
opencurly = tok $ literal '{'

*Main> runParser opencurly "{;abc\ndef"
Just ('{', "def")

*Main> runParser opencurly "(def"
Nothing
\end{verbatim}

Excellent!  The \verb+opencurly+ parser matches an open curly brace, failing
otherwise, and discard trailing whitespace and comments.

It's straightforward to define parsers for the other three brace tokens:
\begin{verbatim}
closecurly = tok $ literal '}'
openparen  = tok $ literal '('
closeparen = tok $ literal ')'
\end{verbatim}

The parser for our last token, Symbol, is more complicated because any number of
alphabetical characters is valid.  We'll use \verb+satisfy+ and a
predicate to check that a character is valid, and \verb+many1+ to 
match one or more valid characters:
\begin{verbatim}
symbol = tok $ many1 char
  where char = satisfy (flip elem (['a' .. 'z'] ++ ['A' .. 'Z']))

@@

*Main> runParser symbol "abc;sdfasdfsa\n  123"
Just ("abc","123")

*Main> runParser symbol "123;sdfasdfsa\n  abc"
Nothing
\end{verbatim}

Note that we again use \verb+tok+ to throw away trailing junk.

@@

Next, we move on to the special forms.  All special forms are delimited with
curly braces, and lambda and define are our special forms, so we have:

\begin{verbatim}
special = opencurly *> (define <|> lambda) <* closecurly
\end{verbatim}

A straightforward translation of define from the grammar produces:
\begin{verbatim}
define =
    check (== "define") symbol   *>
    pure ADefine                <*>
    symbol                      <*>
    form
\end{verbatim}

On to Lambda!  Not only does Lambda have additional syntax with its parameter 
list, but we also need to ensure that the parameter names are unique.  We can 
do that using the \verb+check+ combinator again:
\begin{verbatim}
lambda = 
    check (== "lambda") symbol      *>
    opencurly                       *>
    pure ALambda                   <*>
    check distinct (many0 symbol)  <*>
    (closecurly                     *>
     many1 form)
  where
    distinct names = length names == length (nub names)
\end{verbatim}

If the symbols are distinct, the parameter list subparser will succeed,
whereas if there's a repeated symbol, it will fail.

\subsection{Finishing touches}
All we have left to do are \verb+form+ (which we've already used recursively
to build the previous parsers) and \verb+woof+.

@@

The final parser, \verb+woof+, needs to not only parse all the forms, but also 
make sure that the entire input is consumed.  The end of input check is done 
with the \verb+end+ parser:
\begin{verbatim}
endCheck = switch item
\end{verbatim}

We also need to discard leading comments and whitespace, so the final Woof 
parser is:
\begin{verbatim}
woof = junk *> many0 form <* endCheck
\end{verbatim}

\subsection{Examples}
Let's look at a few examples of our parser in action.  Here are a couple of
successful examples:
\begin{verbatim}
*Main> runParser woof "; \n   {lambda {x y z} a b (c b a)}end"
Just ([ALambda ["x","y","z"] 
               [ASymbol "a",
                ASymbol "b",
                AApp (ASymbol "c") 
                     [ASymbol "b",
                      ASymbol "a"]],
       ASymbol "end"],
      "")

*Main> runParser woof "a b c {define q (f x)}"
Just ([ASymbol "a",
       ASymbol "b",
       ASymbol "c",
       ADefine "q" (AApp (ASymbol "f") [ASymbol "x"])],"")
\end{verbatim}

And a failing one:
\begin{verbatim}
*Main> runParser woof "a b c {define q (f x)},"
Nothing
\end{verbatim}

Note that while the first two parses consume the entire input, while the last 
one fails because it doesn't recognize the trailing comma.

@@

Let's start with an informal spec of different cases of malformed input, along 
with the error messages that should be reported: 
\begin{itemize}
  \item application: missing operator:  \begin{verbatim}()\end{verbatim}
  \item application: missing close parenthesis:  \begin{verbatim}(a b\end{verbatim}
  \item define: missing symbol:  \begin{verbatim}{define (a b c)}\end{verbatim}
  \item define: missing form:  \begin{verbatim}{define a}\end{verbatim}
  \item lambda: missing parameter list:  \begin{verbatim}{lambda (a b c)}\end{verbatim}
  \item lambda: duplicate parameter names:  \begin{verbatim}{lambda {a b a} (c d)}\end{verbatim}
  \item lambda: missing parameter list close curly:  \begin{verbatim}{lambda {a b (c d)}\end{verbatim}
  \item lambda: missing body form:  \begin{verbatim}{lambda {a b}}\end{verbatim}
  \item special form: unable to parse:  \begin{verbatim}{defin x y}\end{verbatim}
  \item special form: missing close curly:  \begin{verbatim}{define x y\end{verbatim}
  \item woof: unparsed input:  \begin{verbatim}a,b\end{verbatim}
\end{itemize}

To allow error reporting in our parsers, we first need to extend our monad stack, 
adding in a layer for errors.  We'll use a second monad transformer type class 
-- MonadError, replace the Maybe layer with its corresponding transformer datatype 
MaybeT, and add in an Either layer on the bottom of the stack.  Our parser stack 
and \verb+runParser+ function are now:
\begin{verbatim}
type Parse e t a = StateT [t] (MaybeT (Either e)) a

@@

*Main> runParser (throwError "no!" <|> satisfy (== 'b')) "babc"
Left "no!"
\end{verbatim}

We'll take advantage of this backtracking restriction to produce accurate error
messages.

@@

\subsection{Error messages}
We must modify the remaining parsers to report meaningful error messages; let's
define the messages that will be reported in each case:
\begin{verbatim}
eAppOper    =  "application: missing operator"
eAppClose   =  "application: missing close parenthesis"
eDefSym     =  "define: missing symbol"
eDefForm    =  "define: missing form"
eLamParam   =  "lambda: missing parameter list"
eLamDupe    =  "lambda: duplicate parameter names"
eLamPClose  =  "lambda: missing parameter list close curly"
eLamBody    =  "lambda: missing body form"
eSpecClose  =  "special form: missing close curly"
eSpecial    =  "special form: unable to parse"
eWoof       =  "woof: unparsed input"
\end{verbatim}

@@

What data do we want to include in our errors?  A String describing the error 
and a token reporting the position will do:
\begin{verbatim}
type Error = String
\end{verbatim}

We'll first add error-reporting to \verb+application+:
\begin{verbatim}
application =
    openparen                     >>
    commit eAppOper form          >>= \op ->
    many0 form                    >>= \args ->
    commit eAppClose closeparen   >>
    return (AApp op args)
\end{verbatim}

@@

The \verb+define+ and \verb+lambda+ parsers are upgraded similarly:
\begin{verbatim}
define =
    check (== "define") symbol    *>
    pure ADefine                 <*>
    commit eDefSym symbol        <*>
    commit eDefForm form

lambda =
    check (== "lambda") symbol       >>
    commit eLamParam opencurly       >>
    many0 symbol                     >>= \params ->
    (if distinct params 
        then return ()
        else throwError eLamDupe)    >>
    commit eLamPClose closecurly     >>
    commit eLamBody (many1 form)     >>= \bodies ->
    return (ALambda params bodies)
  where
    distinct names = length names == length (nub names)
\end{verbatim}

@@

\subsection{Error messages:  finishing touches}
To complete the error-reporting parser, we need to change a couple more details. 
First, we want an error reported when an open curly brace is found, but no 
special form can be parsed.  We extend the \verb+special+ 
parser to report an error if it matches an open brace but not define or lambda:
\begin{verbatim}
special = 
    opencurly  *>  commit eSpecial (define <|> lambda)  <*  commit eSpecClose closecurly
\end{verbatim}

@@

When we can no longer parse any forms, instead of failing, we'd like to report 
an error if there's any input left, and otherwise succeed:
\begin{verbatim}
woof = junk *> many0 form <* commit eWoof endCheck
\end{verbatim}

@@

Knowing that a close-parenthesis is missing somewhere deep inside a nested mess
of forms isn't very helpful -- you also want to know where it happened.  Let's 
add that to our parser.
